---
title: Replicate Paper with R and rCharts
author: Timely Portfolio
github: {user: timelyportfolio, repo: rCharts_factor_analytics, branch: "gh-pages"}
framework: bootplus
layout: post
mode: selfcontained
highlighter: prettify
hitheme: twitter-bootstrap
lead : >
  Fact, Fiction and Momentum Investing
assets:
  jshead:
    - http://d3js.org/d3.v3.min.js
    - http://dimplejs.org/dist/dimple.v1.1.5.min.js
  css:
    - "http://fonts.googleapis.com/css?family=Raleway:300"
    - "http://fonts.googleapis.com/css?family=Oxygen"    
---

# New Working Paper on Momentum

<style>
body{
  font-family: 'Oxygen', sans-serif;
  font-size: 15px;
  line-height: 22px;
}

h1,h2,h3,h4 {
  font-family: 'Raleway', sans-serif;
}

.tooltip{
  opacity:1 !important
}
</style>

```{r echo = F, warning=F, error=F, message=F}
require(knitr)
opts_chunk$set(warning=F, error=F, message=F, fig.width = 10, fig.height = 6, results='asis', cache = F, tidy = F, echo = F)
options(rcharts.mode="iframe")
```

It seems that [Gary Antonacci from Optimal Momentum](http://optimalmomentum.blogspot.com/2014/06/fact-fiction-and-momentum-investing.html) and I were simultaneously enjoy this fine working paper on momentum.

<blockquote>
<strong>Fact, Fiction and Momentum Investing</strong><br>
Asness, Clifford S. and Frazzini, Andrea and Israel, Ronen and Moskowitz, Tobias J.<br>
available at <a href="http://papers.ssrn.com/sol3/papers.cfm?abstract_id=2435323">SSRN</a><br>
May 9, 2014<br>
</blockquote>

I have had it in an open Chrome tab since the day it was posted to SSRN. Even after reading a couple of times, I left it open intending to attempt a replication in R since the data was available, the math was understandable, and the topic was interesing.  Beyond a simple replication, I also wanted to add some [rCharts](http://rcharts.io) and [slidify](http://slidify.io).  Below is a fairly complete replication of myths 1 - 3.  I have chosen `echo = F` to hide most of the code.  If you would like to see the code and replicate for yourself, please see the [Github repository](https://github.com/timelyportfolio/rCharts_factor_analytics/).


---
### Data Source

Once again this amazing resource [Kenneth French Data Library](http://mba.tuck.dartmouth.edu/pages/faculty/ken.french/data_library.html) will be our data source. We will use the monthly data files.  My code to retrieve these is ugly.  Feel free to functionalize it if you have the desire.


```{r eval = F, echo = F}
require(reshape2)
require(quantmod)
require(PerformanceAnalytics)
require(rCharts)
require(Gmisc)

#use monthly French data to replicate research on momentum in R
# data source : Kenneth French Data Library
#               http://mba.tuck.dartmouth.edu/pages/faculty/ken.french/data_library.html 
# paper       : Fact, Fiction and Momentum Investing
#               Asness, Clifford S. and Frazzini, Andrea and Israel, Ronen and Moskowitz, Tobias J.
#               May 9, 2014
#               http://papers.ssrn.com/sol3/papers.cfm?abstract_id=2435323

#gather french factor data
my.url="http://mba.tuck.dartmouth.edu/pages/faculty/ken.french/ftp/F-F_Research_Data_Factors.zip"
my.tempfile<-paste(tempdir(),"\\frenchfactors.zip",sep="")
my.usefile<-paste(tempdir(),"\\F-F_Research_Data_Factors.txt",sep="")
download.file(my.url, my.tempfile, method="auto", 
              quiet = FALSE, mode = "wb",cacheOK = TRUE)
unzip(my.tempfile,exdir=tempdir(),junkpath=TRUE)
#read space delimited text file extracted from zip
french_factors <- read.table(file=my.usefile,
                             header = TRUE, sep = "",
                             as.is = TRUE,
                             skip = 3, nrows=1054)
#get xts for analysis
french_factors_xts <- as.xts(
  french_factors,
  order.by=as.Date(
    paste0(rownames(french_factors),"01"),
    format="%Y%m%d"
  )
)

#now get the momentum factor
my.url="http://mba.tuck.dartmouth.edu/pages/faculty/ken.french/ftp/F-F_Momentum_Factor.zip"
my.usefile<-paste(tempdir(),"\\F-F_Momentum_Factor.txt",sep="")
download.file(my.url, my.tempfile, method="auto", 
              quiet = FALSE, mode = "wb",cacheOK = TRUE)
unzip(my.tempfile,exdir=tempdir(),junkpath=TRUE)
#read space delimited text file extracted from zip
french_momentum <- read.table(file=my.usefile,
                              header = TRUE, sep = "",
                              as.is = TRUE,
                              skip = 12, nrows=1048)
#get xts for analysis
french_momentum_xts <- as.xts(
  french_momentum,
  order.by=as.Date(
    paste(rownames(french_momentum),"01"),
    format="%Y%m%d"
  )
)

#merge UMD (momentum) with other french factors
french_factors_xts <- na.omit( merge( french_factors_xts, french_momentum_xts ) )
french_factors_xts <- french_factors_xts/100
```

---
### Table 1

For Table 1 the authors try to dispel the notion

<blockquote>
Myth #1: Momentum returns are too “small and sporadic”.
</blockquote>

Instead of a table I thought some [rCharts](http://rcharts.io) + [dimplejs](http://dimplejs.org) for an interactive table might help visualize the annualized returns and Sharpe ratios of each of the four factors.

```{r}
#test our numbers
#these are the sample periods
periods <- c("1927::2013","1963::2013","1991::2013")
table1.returns <- t(sapply(
  periods,
  function(x){
    return(data.frame(Return.annualized(french_factors_xts[x,-4],geometric=F)))
  }
))

table1 <- melt(
  data.frame(
    Sample = rownames(table1.returns),
    table1.returns,
    Metric = rep("Returns",nrow(table1.returns))
  ),
  id.vars = c("Sample","Metric"),
  variable.name = "factor"
)

table1.sharpe <- t(do.call(cbind,lapply(
  periods,
  function(x){
    df <- data.frame(
      SharpeRatio.annualized(
        french_factors_xts[x,-4],
        Rf = 0,
        geometric=F
      )[1,]
    )
    colnames(df) <- x
    return(df)
  }
)))

table1 <- rbind(
  table1,
  melt(
    data.frame(
      Sample = rownames(table1.sharpe),
      table1.sharpe,
      Metric = rep("Sharpe",nrow(table1.sharpe))
    ),
    id.vars = c("Sample","Metric"),
    variable.name = "factor"
  )
)

d1 <- dPlot(
  y = "value",
  x = c("Metric","Sample"),
  groups = c("Metric","factor"),
  data = table1,
  type = "line",
  xAxis = list( orderRule = "Sample" ),
  yAxis = list( outputFormat = "0.2%" ),
  defaultColors = "#!d3.scale.category10()!#",
  height = 350,
  width = 750
)
d1$setTemplate(
  script = "./chart_assignColors.html"
)
d1

```

Combining Sharpe and return on the same chart is not ideal, but I thought it would demonstrate some of the power of dimplejs.  A facetted approach or separate (small-multiples) charts here would work much better since the scales are so different.

---
### Table 2

```{r}
#ok Table 1 matches so now let's move on to other calculations
#Table 2
table2.1year <-  do.call(rbind,lapply(
  periods,
  function(x){
    df <- data.frame(lapply(
      rollapply(french_factors_xts[x,-4], width = 12, by = 1, FUN = Return.cumulative, geometric=F),
      function(y){sum(na.omit(y)>=0)/nrow(na.omit(y))}
    ))
    rownames(df) <- x
    return(df)
  }
))

table2 <- melt(
  data.frame(
    Sample = rownames(table2.1year),
    table2.1year
  ),  
  id.vars = "Sample",
  variable.name = "factor"
)
```

For Table 2 which also seeks to refute the "small and sporadic" Myth #1, we can employ the [Gmisc](http://gforge.se/gmisc/) package to easily produce a fairly good looking HTML table.

```{r}
htmlTable(
  round(table2.1year * 100,2),
  align = c('c',rep('r',ncol(table2.1year))),
  title = "Sample",
  caption = "Table 2: Probability of 1 Year Positive Returns"
)
```
<br>
I will not produce a chart for every table, but I feel like a couple more will not hurt anything.  As in the previous chart, this will use rCharts and dimplejs.

```{r}
d2 <- dPlot(
  y = "value",
  x = c("Sample","factor"),
  groups = c("Sample","factor"),
  data = table2,
  type = "bar",
  xAxis = list( orderRule = "Sample" ),
  yAxis = list( outputFormat = ".1%" ),
  defaultColors = "#!d3.scale.ordinal().range(['#196396','#d56a0b','#9d9e1c','#139ead']).domain(['SMB','Mom','HML','Mkt.RF'])!#",
  height= 350
)
d2$setTemplate(
  script = "./chart_assignColors.html"
)
d2
```

Now for the right half of Table 2 which shows the probability of 5 year positive returns, let's do the same thing with a table and then a chart.

```{r}
table2.5year <- do.call(rbind,lapply(
  periods,
  function(x){
    df <- data.frame(lapply(
      rollapply(french_factors_xts[x,-4], width = 60, by = 1, FUN = Return.cumulative, geometric=F),
      function(y){sum(na.omit(y)>=0)/nrow(na.omit(y))}
    ))
    rownames(df) <- x
    return(df)
  }
))

table2 <- melt(
  data.frame(
    Sample = rownames(table2.1year),
    table2.1year
  ),  
  id.vars = "Sample",
  variable.name = "factor"
)

htmlTable(
  round(table2.5year * 100,2),
  align = c('c',rep('r',ncol(table2.5year))),
  title = "Sample",
  caption = "Table 2: Probability of 5 Year Positive Returns"
)
```

<br>
I have been using `iframe` mode with rCharts, but for example purposes, I will show this next chart with `inline`. `inline` can be cleaner but with multiple charts and especially multiple libraries can be problematic with conflicts.

```{r}
d2 <- dPlot(
  y = "value",
  x = c("Sample","factor"),
  groups = c("Sample","factor"),
  data = table2,
  type = "bar",
  xAxis = list( orderRule = "Sample" ),
  yAxis = list( outputFormat = ".1%" ),
  defaultColors = "#!d3.scale.ordinal().range(['#196396','#d56a0b','#9d9e1c','#139ead']).domain(['SMB','Mom','HML','Mkt.RF'])!#",
  height = 350
)
d2$setTemplate(
  script = "./chart_assignColors.html"
)
d2$show("inline")
```


---
### Table 3

Table 3 offers even more proof for the arguments against "small and sporadic".  It blends pieces of Table 2 with the Sharpe Ratio and probability of positive returns on a portfolio of 60% HML (Value) and 40% UMD (Momentum).  This simple portfolio produces very solid results and inspires me to further my exploration of the rebalancing concept introduced in my post [Unsolved Mysteries of Rebalancing](http://timelyportfolio.blogspot.com/2014/02/unsolved-mysteries-of-rebalancing.html).

```{r}
#Table 3

#start table 3 with data from table2
table3 <- data.frame(t(data.frame(
  table1.sharpe[1,],
  t(table2.1year[1,]),
  t(table2.5year[1,])
)))

sr <- as.vector(SharpeRatio.annualized(
  apply(
    french_factors_xts[periods[1],c(3,5)],
    MARGIN = 1,
    function(x){ x[1] * 0.6 + x[2] * 0.4}
  ),
  Rf = 0,
  geometric=F
))

table3$HML60_UMB40 <- c(
  sr,
  sapply(
    c(12,60),
    function(width){
      y = rollapply(
        apply(
          french_factors_xts[,c(3,5)],
          MARGIN = 1,
          function(x){ x[1] * 0.6 + x[2] * 0.4}
        ),
        width = width,
        by = 1,
        FUN = Return.cumulative,
        geometric=F
      )
      return(sum(na.omit(y)>=0)/length(na.omit(y)))
    }
  )
)

htmlTable(
  format(round(table3,2),digits=2),
  align = c('c',rep('c',ncol(table2.5year))),
  title = NULL,
  rowname = c("Sharpe Ratios", "% Positive, 1-year Rolling", "% Positive, 5-year Rolling"),
  caption = "Table 3: 1927 - 2013"
)
```

---
### More French Data

We will need to pull in some more French data to work through the next set of tables.  This data [6 Portfolios Formed Monthly on Size (3) x Momentum (2)](http://mba.tuck.dartmouth.edu/pages/faculty/ken.french/Data_Library/det_6_port_form_sz_pr_12_2.html) will allow us to decompose the momentum factor into short and long buckets and also small and large size buckets.

```{r eval = F}
#Table 4
#need some additional data here
#http://mba.tuck.dartmouth.edu/pages/faculty/ken.french/ftp/6_Portfolios_ME_Prior_12_2.zip
my.url="http://mba.tuck.dartmouth.edu/pages/faculty/ken.french/ftp/6_Portfolios_ME_Prior_12_2.zip"
my.tempfile<-paste(tempdir(),"\\frenchfactors.zip",sep="")
my.usefile<-paste(tempdir(),"\\6_Portfolios_ME_Prior_12_2.txt",sep="")
download.file(my.url, my.tempfile, method="auto", 
              quiet = FALSE, mode = "wb",cacheOK = TRUE)
unzip(my.tempfile,exdir=tempdir(),junkpath=TRUE)
#read space delimited text file extracted from zip
french_hml_smb <- read.table(file=my.usefile,
                             header = TRUE, sep = "",
                             as.is = TRUE,
                             skip = 12, nrows=1048)
colnames(french_hml_smb) <- c(
  paste0(
    "Small.",
    colnames(french_hml_smb)[1:3]
  ),
  paste0(
    "Big.",
    colnames(french_hml_smb)[1:3]
  )
)
#get xts for analysis
french_hml_smb_xts <- as.xts(
  french_hml_smb/100,
  order.by=as.Date(
    paste0(rownames(french_hml_smb),"01"),
    format="%Y%m%d"
  )
)

#so short side will be average of Small.Low and Big.Low
#and long side will be the average of Small.High and Big.High
french_short_long_xts <- as.xts(do.call(rbind,apply(
  french_hml_smb_xts,
  MARGIN=1,
  function(period){
    data.frame(
      shortside = mean(period[c(1,4)]),
      longside = mean(period[c(3,6)])
    )
  }
)),order.by = index(french_hml_smb_xts))

#add umd from the factors
#as check this should be same as sum of -short and long
french_short_long_xts <- merge(
  french_short_long_xts,
  french_factors_xts[,c(5,1,4)] #Market and Mom (UMD)
)
```

---
### Table 4

Now that we have the additional French data, we can replicate the authors` attempt to disprove

<blockquote>
Myth #2: Momentum cannot be captured by long-only investors as “momentum can only be exploited on the short side”.
</blockquote>

My numbers differ slightly on the market-adjusted returns (left half), but are close enough to think that the approach is the same.


```{r}
#UMD market-adjusted returns (alpha)
table4.lefthalf <- do.call(rbind,lapply(
  periods,
  function(period){
    df <- data.frame(
      shortside = -((CAPM.alpha(
        Ra = french_short_long_xts[period,]$shortside,
        Rb = french_short_long_xts[period,]$Mkt.RF + french_short_long_xts[period,]$RF,
        Rf = french_short_long_xts[period,]$RF
      )+1)^12-1),
      longside = (CAPM.alpha(
        Ra = french_short_long_xts[period,]$longside,
        Rb = french_short_long_xts[period,]$Mkt.RF + french_short_long_xts[period,]$RF,
        Rf = french_short_long_xts[period,]$RF
      )+1)^12-1
    )
    df$UMD <- df$shortside + df$longside
    rownames(df) <- period
    return(df)
  }
))

#UMD returns minus market
table4.righthalf <- do.call(rbind,lapply(
  periods,
  function(period){
    df <- data.frame(
      shortside = -Return.annualized(
        french_short_long_xts[period,]$shortside - 
          french_short_long_xts[period,]$Mkt.RF - 
          french_short_long_xts[period,]$RF,
        geometric=F
      ),
      longside = Return.annualized(
        french_short_long_xts[period,]$longside - 
          french_short_long_xts[period,]$Mkt.RF - 
          french_short_long_xts[period,]$RF,
        geometric=F
      )
    )
    df$UMD <- df$shortside + df$longside
    rownames(df) <- period
    return(df)
  }
))

table4 <- cbind(
  table4.lefthalf,
  table4.lefthalf$longside/table4.lefthalf$UMD,
  table4.righthalf,
  table4.righthalf$longside/table4.righthalf$UMD
)

htmlTable(
  format(round(table4 * 100, 2),digits=3),
  caption = "Table 4",
  title = "Sample",
  align = rep("c", ncol(table4) + 1),
  headings = rep(c("Short Side","Long Side","UMD", "% Long"),2),
  cgroup = c("UMD market-adjusted returns","UMD returns minus market"),
  n.cgroup = c(4,4)
)

```

---
### Table 5

The last myth that we will tackle here, but only the 3rd of 10 myths in the paper, is

<blockquote>
Myth #3: Momentum is much stronger among small cap stocks than large caps.
</blockquote>

We will use the same French data (HML (3) x SMB (2)) from Table 4 to replicate Table 5.  I will leave the right half (Value) of this table as homework.


```{r}
#table 5
table5 <- do.call(rbind,lapply(
  periods,
  function(period){
    df <- data.frame(Return.annualized(
      french_hml_smb_xts[period,],
      geometric = F
    ))
    rownames(df) <- period
    return(df)
  }
))
#umd small
umdsmall <- data.frame(table5$Small.High - table5$Small.Low)
rownames(umdsmall) <- periods
colnames(umdsmall) <- "UMD Small"

#umd big
umdbig <- data.frame(table5$Big.High - table5$Big.Low)
rownames(umdbig) <- periods
colnames(umdbig) <- "UMD Big"

table5 <- cbind(
  umdsmall,
  umdbig,
  data.frame(table1.returns)[,4,drop=F]
)

table5[,3] <- as.numeric(table5[,3])

htmlTable(
  format(round(table5 * 100, 2),digits = 3),
  caption = "Table 5",
  title = "Sample",
  align = rep("c", ncol(table4) + 1),
  cgroup = c("Momentum"),
  n.cgroup = ncol(table5)
)

#by this point I hope you can do the Value piece of table 5 on your own
#if you really can't figure it out, let me know
```

---
### Thanks
As I hope you can tell, this post was more a function of the efforts of others than of my own.

Thanks specifically:
- [Kenneth French](http://mba.tuck.dartmouth.edu/pages/faculty/ken.french/) for his very generous [data library](http://mba.tuck.dartmouth.edu/pages/faculty/ken.french/data_library.html)
- [the AQR research factory](http://www.aqr.com/) for this and all their other research
- [Ramnath Vaidyanathan](http://ramnathv.github.io/) for [rCharts](http://rcharts.io/site) and [slidify](http://slidify.org).
- [John Kiernander](https://twitter.com/jkiernander) for [dimplejs](http://dimplejs.org)
- Nameless Fixed Income Shop for the original chart.
- [Mike Bostock](http://bost.ocks.org/mike/) for everything.
- [Marcello Palmitessa](http://aozora.github.io/bootplus/) for the Bootplus framework.
- Google fonts [Raleway](http://www.google.com/fonts/specimen/Raleway) and [Oxygen](http://www.google.com/fonts/specimen/Oxygen)
